Historical advances in medicine that did not need (or want) a randomized clinical

### 2020-01-26 paper/paper.Rmd
Finished first draft of paper

---
title: Historical advances in medicine that did not need (or want) a randomized clinical
  trial.
author: "Steve Simon"
nocite: |
  @zika-overview, @zika-case-series
output:
  word_document: default
  html_document: default
bibliography: reference.bib
---

@@

Understanding when randomized clinical trials (RCTs) are needed comes from first knowing when those RCTs are not needed. There are many settings where major improvements in health care came through the use of observational studies. Observational studies have some strengths, such as greater ecologic validity and the ability to study larger populations over longer periods of time, but they have many weaknesses and require much more effort to eliminate other possible causal explanation. Some examples of the triumphs of observational research are establishment of a causal association between cigarette smoking and lung cancer, identification of most of the well-known risk factors for heart disease, and discovery of the link between aspirin and Reye’s syndrome. The work was not easy; but the careful and methodical approach offers a lesson in how to assemble information from a series of less than perfect observational studies to produce a definitive conclusion. These examples help to teach students to recognize when the ethical and logistical challenges associated with RCTs are so great that observational studies, with all their limitations, are preferable.

@@

What was lacking in all the research studies utilized in the war against Zika, was the randomized control trial (RCT). An informal Pubmed search on the terms “Zika” and “microcephaly” yielded 1,716 publications, but when you click on the link to the right that limits your search to clinical trials, the list dwindles to a single citation that on closer examination is actually a cohort study.

It is not at all surprising that no RCTs were utilized in this investigation. What is surprising, at least to some, is how much was learned so quickly utilizing a range of research methodologies that are commonly held to be inferior to the RCT. In a few short years, researchers were able to track the spread of the Zika virus, identify the modes of transmission, provide a causal link between Zika infection during pregnancy and microcephaly, and characterize in great detail the impact of this condition on early childhood development.

If you want to understand the benefits of RCTs and when they are needed, you should start by examining the times that they have not been needed. This paper will outline three areas where major progress was made without the benefit of RCTs and discuss what this means for the use of RCTs in research.

@@

>>>>

---
title: Historical advances in medicine that did not need (or want) a randomized clinical
  trial.
author: "Steve Simon"
output:
  word_document: default
  html_document: default
bibliography: reference.bib
csl: american-medical-association.csl
---

@@

Understanding when randomized clinical trials (RCTs) are needed comes from first knowing when those RCTs are not needed. There are many settings where major improvements in health care came through the use of observational studies. Observational studies have some strengths, such as greater ecologic validity and the ability to study larger populations over longer periods of time, but they have many weaknesses and require much more effort to eliminate other possible causal explanation. Some examples of the triumphs of observational research are establishment of a causal association between cigarette smoking and lung cancer, identification of most of the well-known risk factors for heart disease, and discovery of the link between aspirin and Reye’s syndrome. The work was not easy; but the careful and methodical approach offers a lesson in how to assemble information from a series of less than perfect observational studies to produce a definitive conclusion. Understanding the circumstances where observational studies shine will help to teach students to take a balanced approach to research. At the same time, defining the times when where RCTs are not needed will help define better by negation when RCTs are needed.

@@

What was lacking in all the research studies utilized in the war against Zika, was the randomized control trial (RCT). It is not at all surprising that no RCTs were utilized in this investigation. What is surprising, at least to some, is how much was learned so quickly utilizing a range of research methodologies that are commonly held to be inferior to the RCT. In a few short years, researchers were able to track the spread of the Zika virus, identify the modes of transmission, provide a causal link between Zika infection during pregnancy and microcephaly, and characterize in great detail the impact of this condition on early childhood development.

If you want to understand the benefits of RCTs and when they are needed, you should start by examining the times that they have not been needed. There are wonderful historical reviews of the research that first drew a link between smoking and lung cancer (@statistics-in-action, @report-anniversary), that established major risk factors for heart disease (@framingham-epidemiology, @framingham-legacy), and that identified aspirin use during viral infections as the predominant cause of Reye's syndrome (@reyes-history). This paper will outline three areas how these research areas relied on a broad range of observational studies and made major progress was made without the benefit of RCTs. This review has broad implications for when RCTs should be used in research.

@@

Around the same time, several case control studies were being published. The results of one study (@reyes-case-control-2) are representative of the general findings. The researchers identified 159 cases of Reye's syndrome following an illness like influenze or varicella. Recruiting was done at six pediatric centers in Ohio between December 1978 and March 1980. Controls were recruited from the same school and matched on age, sex, and race and had an antecedant illness within one week of the case. The use of salicylates (aspirin) among the cases was significantly higher (97% versus 71%) among the cases.

While the case control studies alone may not have provided a definitive link between aspirin and Reye's syndrome, it was sufficiently persuasive to start a publicity effort warning the public about the possible danger associated with aspirin use in a child during a viral illness. What happened next was definitive proof. The number of cases reported to the CDC surveillance system dropped steadily from a high of 555 cases in 1980 to no more than 2 per year in 1994-1997 and the drops exhibited a strong temporal association with public health campaign events (@reyes-history, Figure 1).

The near total elimination of Reye's syndrome has been properly hailed as a public health triumph (@reyes-disappearance) and was accomplished without any RCTs. The humble case-control study combined with temporal trends in disease surveillance were all that was needed.

#### Discussion

The recent research effort to examine the link between the Zika virus and microcephaly mirrors past research efforts in lung cancer, heart disease, and Reye's syndrome. In every case, major advances in knowledge were accomplished using only observational studies.

The effective use of observational studies is not limited to just these areas. Benjamin Djulbegovic has developed an web page inventory of non-randomized trials that have changed clinical practice (http://personal.health.usf.edu/bdjulbeg/oncology/NON-RCT-practice-change.htm) that covers studies in Endocrinology, Infectious Diseases, Oncology, Hematology, Pulmonology, Urology, Cardiology, Musculoskeletal conditions, Psychiatry, and Surgery.

The research efforts described in this paper share some common features that can help us understand when RCTs are not needed. The most obvious feature is the practical and ethical barriers to randomization. No one could envision a study where some children with a viral fever are given aspirin to see if any harm could come from it. The long latency periods associated with lung cancer and heart disease would saddle most RCTs with an unworkable time frame.

Second, these observational studies represented pioneering efforts in an area where little is known. This is illustrated best, perhaps, by the Framingham study. The original planners of this study had considered using the population of Framingham, Massachusetts as a testbed for trying new interventions, but rejected that approach in favor of an observational study because there weren't any interventions at the time that warranted testing (@framingham-epidemiology). Today there are many RCTs testing a variety of prevention initiatives, but these are only possible because of the foundational knowledge that Framingham provided.

Third, these studies were looking to attack problems by taking big bites rather than nibbling around the edges. The odds ratios in the lung cancer studies were showing a 10 to 20 fold increase in risk among smokers. The finding of aspirin use in the medical history of almost every Reye's syndrome case was a veritable smoking gun and made any concerns about biases in a case-control study seem unimportant.

A final commonality among these studies is the hard work that researchers put in place to insure that these observational studies would produce results that could withstand critical scrutiny. Elaborate matching plans helped to insure a comparable control group and efforts to use blinding during patient interviews helped to reduce biases from researcher expectations. The recent studies of the Zika virus show that same level of care. One example is the careful consideration of what specific head circumference measurements would be considered as evidence of microcephaly (@zika-interpret).

It's been said many times how much easier the conduct of an RCT is. Get a bunch of patients together, flip a coin to see who gets the drug and who gets the placebo, wait three years and then count the number of dead bodies. Watching the care and concern of the researchers in these observational studies makes one hope that some of this could rub off on the researchers running RCTs. While RCTs may have fewer and less severe biases than observational studies, they do indeed have biases.

So let's flip the coin over to the other side. If observational studies are what we want when studying the link between smoking and lung cancer, the risk factors for heart disease, the cause of Reye's syndrome, and the microcephaly associated with the Zika virus, then when do we want to use RCTs instead?

RCTs do very well in a mature setting where we know a lot about the causes of disease and can develop some well supported interventions. They are great at developing and testing a series of small incremental improvements that nibble around the edges. If you do enough nibbling, you can make some big improvements over time.

The RCT excels when you are in the muddled middle. You have some interventions that sort of work but there's hope that you might find something a bit better. When comparing a fair intervention to a good intervention, biases become very troublesome, and the ability to remove many of these biases through randomization helps to insure that you'll recognize the distinction between fair and good.

I once worked with a statistician who told me that he would never work on a study that didn't use randomization. My first thought was "You have some many clients asking for your help that you can send the half away that come to you with observational studies?" But I later came to realize that this was a type of arrogance. We can have debates about how observational studies have greater ecologic validity or how randomized studies control for many biases but this is really not the point. The type of problem you are studying should determine what type of research you choose. You don't want to be the proverbial worker who sees nothing but nails because the only tool you have is a hammer.

#### Conclusion

Research efforts in several major areas have not used RCTs. A review of how these developed important insights and make great progress helps you to understand the settings where observational studies are needed and RCTs are not. This also helps you understand where RCTs are truly needed.
